Example2 gene set analysis: The case of Azacytidine

### 2021-08-25 QC_analysis.Rmd
minor updates

res <- sapply(xyunique,function(myid){
  myqc <- x[which(x$Pubmed.Central.ID==myid),,drop=TRUE]
  ydat <- y[which(y$Pubmed.Central.ID==myid),,drop=TRUE]
  qcres <- sapply(5:16, function(m) {
    mqc <- myqc[[m]]
    mqc <- unlist(strsplit(mqc,", "))
    myy <- ydat[[m]]
    myy <- unlist(strsplit(myy,", "))
  length(intersect(mqc,myy)) / length(union(mqc,myy))
  })
  return(qcres)
})

>>>>

res <- sapply(xyunique,function(myid){
  myqc <- x[which(x$Pubmed.Central.ID==myid),,drop=TRUE]
  ydat <- y[which(y$Pubmed.Central.ID==myid),,drop=TRUE]
  qcres <- sapply(5:16, function(m) {
    mqc <- myqc[[m]]
    mqc <- unlist(strsplit(mqc,", "))
    mqc <- gsub("Not stated","No",mqc)
    myy <- ydat[[m]]
    myy <- unlist(strsplit(myy,", "))
    myy <- gsub("Not stated","No",myy)
  length(intersect(mqc,myy)) / length(union(mqc,myy))
  })
  return(qcres)
})

### 2021-08-25 example2.Rmd
minor updates

---
title: "Example gene set analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

I'm using some RNA-seq data looking at the effect of hyperglycemia on hepatocytes.

@@

mdat<-getDEE2Metadata("hsapiens")
samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
samplesheet$HG<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
s1 <- subset(samplesheet,VPA==0)

@@

# counts 
x1<-x[,which(colnames(x) %in% s1$SRR_accession)]

@@

# MDS plot
plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$HG, main="MDS")

#run DESeq2 for LG vs HG
y <- DESeqDataSetFromMatrix(countData = round(x1), colData = s1, design = ~ HG)
y <- DESeq(y)
LGvHG <- results(y)
LGvHG<-as.data.frame(LGvHG[order(LGvHG$pvalue),])
rownames(LGvHG)<-sapply(strsplit(rownames(LGvHG),"\\."),"[[",1)
head(LGvHG)

@@

maplot(LGvHG,"LG v HG")
make_volcano(LGvHG,"LG v HG")

@@

m <- mitch_import(LGvHG,DEtype = "DEseq2", geneTable = gt)
head(m)
mres <- mitch_calc(m,genesets = genesets)
head(mres$enrichment_result) 
m_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
m_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
length(m_up)
length(m_dn)

@@

LGvHG_up <- rownames(subset(LGvHG,log2FoldChange>0,padj<0.05))
LGvHG_up <- unique(gt[which(rownames(gt) %in% LGvHG_up),1])
head(LGvHG_up)

LGvHG_dn <- rownames(subset(LGvHG,log2FoldChange<0,padj<0.05))
LGvHG_dn <- unique(gt[which(rownames(gt) %in% LGvHG_dn),1])
head(LGvHG_dn)

LGvHG_bg <- rownames(LGvHG)
LGvHG_bg <- unique(gt[which(rownames(gt) %in% LGvHG_bg),1])
length(LGvHG_bg)
head(LGvHG_bg)

c_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
c_up <- head(c_up,1000)
head(c_up)
c_up <- rownames(subset(c_up, p.adjust < 0.05))

c_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
c_dn <- head(c_dn,1000)
head(c_dn)
c_dn <- rownames(subset(c_dn, p.adjust < 0.05))

@@

LGvHG_de <- rownames(subset(LGvHG,padj<0.05))
LGvHG_de <- unique(gt[which(rownames(gt) %in% LGvHG_de),1])
head(LGvHG_de)

d_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
d_de <- head(d_de,1000)
d_de <- rownames(subset(d_de, p.adjust < 0.05))
head(d_de)

@@

LGvHG_bg <- w$GeneInfo$GeneSymbol
length(LGvHG_bg)
head(LGvHG_bg)

f_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
f_up <- head(f_up,1000)
head(f_up)
f_up <- rownames(subset(f_up, p.adjust < 0.05))

f_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, maxGSSize = 5000, TERM2GENE = genesets2)
f_dn <- head(f_dn,1000)
head(f_dn)
f_dn <- rownames(subset(f_dn, p.adjust < 0.05))

@@

e_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, maxGSSize = 5000, TERM2GENE = genesets2)
e_de <- head(e_de,1000)
head(e_de)
e_de <- rownames(subset(e_de, p.adjust < 0.05))

@@

png("images/fcs_ora1.png")
plot(euler(v1),quantities = TRUE, main="FCS vs ORA")
dev.off()

png("images/orabg1.png")
plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")
dev.off()

png("images/oracomb1.png")
plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
dev.off()

>>>>

---
title: "Example2 gene set analysis: The case of Azacytidine"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

I'm using some RNA-seq data looking at the effect of Axacytidine on AML3 cells.

@@

mdat<-getDEE2Metadata("hsapiens")
samplesheet <- mdat[grep("SRP038101",mdat$SRP_accession),]
samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
samplesheet$Aza<-as.factor(c(1,1,1,0,0,0))

@@

# counts 
x1<-x[,which(colnames(x) %in% samplesheet$SRR_accession)]

@@

# MDS plot
plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=samplesheet$Aza, main="MDS")

#run DESeq2 for LG vs HG
y <- DESeqDataSetFromMatrix(countData = round(x1), colData = samplesheet, design = ~ Aza)
y <- DESeq(y)
CTLvAZA <- results(y)
CTLvAZA<-as.data.frame(CTLvAZA[order(CTLvAZA$pvalue),])
rownames(CTLvAZA)<-sapply(strsplit(rownames(CTLvAZA),"\\."),"[[",1)
head(CTLvAZA)

@@

maplot(CTLvAZA,"Ctl v Aza")
make_volcano(CTLvAZA,"Ctl v Aza")

@@

m <- mitch_import(CTLvAZA,DEtype = "DEseq2", geneTable = gt)
head(m)
mres <- mitch_calc(m,genesets = genesets)
head(mres$enrichment_result) 
m_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
m_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
length(m_up)
length(m_dn)

@@

CTLvAZA_up <- rownames(subset(CTLvAZA,log2FoldChange>0,padj<0.05))
CTLvAZA_up <- unique(gt[which(rownames(gt) %in% CTLvAZA_up),1])
head(CTLvAZA_up)

CTLvAZA_dn <- rownames(subset(CTLvAZA,log2FoldChange<0,padj<0.05))
CTLvAZA_dn <- unique(gt[which(rownames(gt) %in% CTLvAZA_dn),1])
head(CTLvAZA_dn)

CTLvAZA_bg <- rownames(CTLvAZA)
CTLvAZA_bg <- unique(gt[which(rownames(gt) %in% CTLvAZA_bg),1])
length(CTLvAZA_bg)
head(CTLvAZA_bg)

c_up <- enricher(gene = CTLvAZA_up, universe = CTLvAZA_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
c_up <- head(c_up,1000)
head(c_up)
c_up <- rownames(subset(c_up, p.adjust < 0.05))

c_dn <- enricher(gene = CTLvAZA_dn, universe = CTLvAZA_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
c_dn <- head(c_dn,1000)
head(c_dn)
c_dn <- rownames(subset(c_dn, p.adjust < 0.05))

@@

CTLvAZA_de <- rownames(subset(CTLvAZA,padj<0.05))
CTLvAZA_de <- unique(gt[which(rownames(gt) %in% CTLvAZA_de),1])
head(CTLvAZA_de)

d_de <- enricher(gene = CTLvAZA_de, universe = CTLvAZA_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
d_de <- head(d_de,1000)
d_de <- rownames(subset(d_de, p.adjust < 0.05))
head(d_de)

@@

CTLvAZA_bg <- w$GeneInfo$GeneSymbol
length(CTLvAZA_bg)
head(CTLvAZA_bg)

f_up <- enricher(gene = CTLvAZA_up, universe = CTLvAZA_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
f_up <- head(f_up,1000)
head(f_up)
f_up <- rownames(subset(f_up, p.adjust < 0.05))

f_dn <- enricher(gene = CTLvAZA_dn, universe = CTLvAZA_bg, maxGSSize = 5000, TERM2GENE = genesets2)
f_dn <- head(f_dn,1000)
head(f_dn)
f_dn <- rownames(subset(f_dn, p.adjust < 0.05))

@@

e_de <- enricher(gene = CTLvAZA_de, universe = CTLvAZA_bg, maxGSSize = 5000, TERM2GENE = genesets2)
e_de <- head(e_de,1000)
head(e_de)
e_de <- rownames(subset(e_de, p.adjust < 0.05))

@@

png("images/fcs_ora2.png")
plot(euler(v1),quantities = TRUE, main="FCS vs ORA")
dev.off()

png("images/orabg2.png")
plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")
dev.off()

png("images/oracomb2.png")
plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
dev.off()
